---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data


```{r}
url <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
```

Data for this project comes from `r url`

Download the data, and capture and report metadata


```{r}
zipFile <-"activity.zip"
dataFile <- "activity.csv"
dataInfoFile <- "dataSourceInformation.csv"

    #Lectures say to download data from the original source. The assigment says
    #to use the data in the repo. I am going with what the assignement says,
    #however, if one were to follow the lectures, uncomment the next line.
#download.file(url,zipFile)
zipInfo <- unzip(zipFile,list=TRUE)
    #Expecting a single csv file, abort if that is not the case
stopifnot(nrow(zipInfo)==1)
stopifnot(zipInfo[1,"Name"]==dataFile)
    #extract the data file, and confirm it is what we are expecting.
unzip(zipFile)
stopifnot(file.exists(dataFile))
    #Create record of data source. Again, good practice per lectures, but not
    #part of the assignement.
    #todo: structure this as rows instead of columns. Time permitting.
write.csv( data.frame(
                Source=url
              , Download_Date = as.POSIXlt(file.mtime(zipFile),"UTC")
              , File_Date = zipInfo[1,"Date"])
        , file = dataInfoFile
        , row.names = FALSE )
print(read.csv( dataInfoFile ))
activity <- read.csv(dataFile)
```


## What is mean total number of steps taken per day?


```{r}
steps<-aggregate(steps ~ date, activity, sum)[,2]
meanSteps <- mean(steps)
medianSteps <- median(steps)
```

Steps Per Day

- Mean:   `r sprintf("%2.2f", meanSteps)`  
- Median:  `r sprintf("%2.2f", medianSteps)`



```{r}
hist(steps,breaks=10,main="Histogram of steps per day")
```



## What is the average daily activity pattern?

```{r}
    # Get steps averaged by time interval
steps <- aggregate(steps ~ interval, activity, mean, rm.na=TRUE)
    # Make time more readable. Use 24 hour hh:mm format.
steps$time <- sprintf("%02d:%02d", steps$interval / 100, steps$interval %% 100)
    # sort by steps and extract the top ten.
topStepTimes <- steps[order(0.0-steps$steps),3:2][1:10,]
```

The time interval with most steps is `r topStepTimes[1,1]`, with an average of `r topStepTimes[1,2]` steps per day.

All of the ten intervals with the most average steps were near this time.

```{r}
print(topStepTimes, row.names = FALSE)
```

The step distribution through the day is shown in this graph

```{r}
plot(steps[,1:2]
     , type="l"
     , main="Average steps per five minute intervals"
     , xlab="Hour of the day"
     , xaxt="n")
    # Replace the time format used in the dataset by the hour of the day on the x axis.
    # Put ticks for even numbered hours
axis(1, at=0:12*200, labels=0:12*2)
```



## Imputing missing values
```{r}
naCount <- sum(is.na(activity$steps))
naPercent <- 100 * naCount / nrow(activity)
```

The source data has `r naCount` missing data points. This represents `r sprintf( "%2.2f", naPercent)`% of the values

We are going to replace the missing data points with the average for that time interval across all days.


```{r}
    #compute mean steps by interval
meanStepsByInterval <- aggregate(steps ~ interval, activity, mean, rm.na=TRUE)
names(meanStepsByInterval) <- c("interval", "mean")
    #add a column for the mean
adjustedData <- merge(activity,meanStepsByInterval,by.x="interval",by.y="interval")
    #add a column, if na then use the mean, otherwise the actual step count
adjustedData$steps.adjusted <- ifelse(is.na(adjustedData$steps), adjustedData$mean, adjustedData$steps)
```

The first few rows of the imputed dataset
```{r}
adjustedData[1:5,]
```

```{r}
    #sum by dates, then drop down to a single column of steps.
steps<-aggregate(steps.adjusted ~ date, adjustedData, sum)[,2]
meanSteps2 <- mean(steps)
medianSteps2 <- median(steps)
```

Steps Per Day

- Mean:   `r sprintf("%2.2f", meanSteps2)`  
- Median:  `r sprintf("%2.2f", medianSteps2)`

```{r}
meanDif <- 100 * (meanSteps2 - meanSteps) / meanSteps
medianDif <- 100 * (medianSteps2 - medianSteps) / medianSteps
```

These values different from the original values by `r sprintf("%2.2f", meanDif)`% and 
`r sprintf("%2.2f", medianDif)`% respectively


```{r}
hist(steps,breaks=10,main="Histogram of adjusted steps per day")
```


## Are there differences in activity patterns between weekdays and weekends?
